Sampling Bayesian probabilities given only sampled priors

### 2025-01-20 main.tex
Some proofing and text for application.

Our motivating situation is that we have a vector of observable summary statistics \vecc, such as the binned 2-point correlation functions of cosmic fields, that we are using to constrain a set of parameters of interest \vecq, which are cosmological parameters such as $\Omega_m, \sigma_8,$ etc.  There is a model $\vecc(\vecq,\vecn)$ for the observables which involves the parameters of interest, but also a vector \vecn\ of nuisance parameters, which include the coefficients of some linear expansion of the redshift distributions $n(z)$ of some population of the galaxies being observed:
\begin{equation}
  n(z) = \sum_{k=1}^{M} n_k b_k(z).
  \label{eq:nzbasis}
\end{equation}
The $b_k$ are a set of predetermined basis functions for the redshift distribution.  In our case there are multiple galaxy populations to be characterized, leading to hundreds of parameters $n_k$ to be considered.

We wish to characterized the Bayesian posterior probability
\begin{equation}
  p(\vecq | \vecc) \propto \int dn\, \likeli(\vecc | \vecq, \vecn) p(\vecq) p(\vecn),
\label{eq:posterior}
\end{equation}
where $\likeli(\vecc | \vecq, \vecn)$ is a known likelihood function of the data, and $p(\vecq)$ and $p(\vecn)$ are priors on the parameters.  This posterior is complex enough that it requires approximation by the output of a Markov chain wandering across the space $(\vecq,\vecn).$

@@

For the DES Y3 analyses, \citet{hyperrank} devised a scheme whereby
the samples of \vecn\ are placed in a grid within some $M$-dimensional
unit hypercube $\mathcal{H}$.  The coordinates \vecu\ within the hypercube
are considered the compressed parameters of $n(z),$ and the
decompression function
$\hat{\vecn}(\vecu)$ outputs the $\vecn_\alpha$ sample at the nearest grid point to
  \vecu.  This solves the problem of creating a continuous \vecu\
  domain, and maintains the equal prior probability of each \vecn\
  sample, but the function \emph{output,} and the resultant likelihood
  function of \vecu\, are discontinuous.
Various strategies are proposed to assign the $\vecn_\alpha$ to the grid points in
$\mathcal{H}$ based on summary statistics, to reduce the
discontinuities.  But the function is never smooth.
As a consequence, many MC samplers become quite inefficient in
sampling of the cosmological posterior.  In particular, samplers such as \textsc{MultiNest} that assume continuity are rendered nearly non-functional.  As a result, the Y3 cosmological priors could not be evaluated with this method.  Instead, the \vecn\ samples were not used, and an \textit{ad hoc} $p(\vecn)$ was adopted which allowed only shifts and dilations of the mean $n(z)$ of the \vecn\ samples.

@@

\section{Derivation}\label{sec:deriv}
We assume that we do have a multivariate normal likelihood for the
observables \vecc\, with the mean being some model
$\hatc(\vecq,\vecn)$ and a fixed covariance matrix $\covm_c.$ In this
section we will assume that the \vecn\ vectors have been shifted by
the mean $\vecn_0 \equiv \left\langle \vecn_\alpha \right\rangle$ so
that their mean is zero.
We seek some function $\hat{\vecn}(\vecu)$ of a lower-dimensional
vector \vecu\ which can be substituted for \vecn\ and yield nearly the
same likelihood function for any \vecn\ in the domain spanned by the
samples $\vecn_\alpha,\; \alpha\in 1\ldots N_\alpha,$ that we are
provided.  This means we want a map $\vecn_\alpha\rightarrow
\vecu_\alpha \rightarrow \hat{\vecn}_\alpha$ such that replacing $\vecn$ with
$\hat\vecn$ alters the cosmological inference by much less than the
other uncertainties.  We will implement this by minimizing the
distance in the space $\vecc$ between the
model generated by $\vecn$ and that by $\hat\vecn$, using the
observations' covariance matrix $\covm_c$ as a metric for the
distance.  This is equivalent to the $\chi^2$ of the difference
between the original and compressed models for the data:
\begin{equation} \left\langle \chi^2 \right\rangle
=  \frac{1}{N_\alpha} \sum_\alpha
                                            \left[ \hatc(\vecq,\vecn_\alpha) - \hatc(\vecq,\hat{\vecn}_\alpha) \right]^T
                                            \covm_c^{-1}
                                            \left[ \hatc(\vecq,\vecn_\alpha) - \hatc(\vecq,\hat{\vecn}\alpha) \right].
\label{eq:chisq}
\end{equation}
If the data are in fact drawn from the model $\hatc(\vecq,\vecn)$ with
a Gaussian likelihood, then this is also the mean shift in $-2\log\likeli$
from the compression.  It is \emph{not}, however, equal to the mean
shift of the overall log of the posterior in
\eqq{eq:posterior}---$\left\langle\chi^2\right\rangle$ is serving as a
proxy for this.

We next assume that the compression is linear, $\hat\vecn=\matX\vecn,$
for some matrix that is idempotent ($\matX\matX = \matX$).  If this is
true, then the contribution to \eqq{eq:chisq} at lowest order in
$\vecn-\hat\vecn$ is
\begin{equation}
  \left\langle \chi^2 \right\rangle = \frac{1}{N_\alpha} \sum_\alpha
  \left[ (\matI-\matX)\vecn_\alpha\right]^T \matF  \left[ (\matI-\matX)\vecn_\alpha\right]
  \label{eq:linearized}
\end{equation}
where we use the Jacobian matrix of the model $\hatc$ to define
\begin{equation}
  \matF \equiv
  \left[\frac{\partial\hatc}{\partial\vecn}\right]_{\vecq_0, \vecn_0}
  \covm_c^{-1} \left[\frac{\partial\hatc}{\partial\vecn}\right]_{\vecq_0,
    \vecn_0}^T.
\label{eq:fisher}
\end{equation}
This quantity is also the Fisher matrix over $\vecn$ giving the information
provided by the observations $\vecc.$  In many cases this matrix will be
rank-deficient and/or poorly conditioned, since the observables are not
likely to be very informative on $\vecn$---if they were, we might not be
concerned with establishing a prior on $\vecn$ to begin with.  We will
not, however, require its inverse.

@@

Transforming the solution back into the space of $\vecn$ yields
\begin{align}
  \matX & = \matT^{-1} \matV_G \proj_M \matV_G^T \matT \\
   & = \left[ \matV_n^T \matLam_n^{1/2} \matV_G \proj_M \right] \left[
     \proj_M \matV_G^T \matLam_n^{-1/2} \matV_n^T \right] \\
\label{eq:DE}
        & \equiv \matD \matE.
\end{align}
We thus obtain our optimal encoding/compression using the nonzero rows of matrix
$\matE$ to give
\begin{equation}
  \vecu_\alpha = \matE \vecn_\alpha
  \label{eq:uu}
\end{equation}
and the decoding/reconstruction of the systematic variables as
\begin{equation}
  \hat\vecn_\alpha = \matD \vecu_\alpha.
  \label{eq:UU}
\end{equation}
One can confirm that this procedure yields a compressed representation
$\vecu$ such that $\covm_u = \matI_M,$ the $M$-dimensional
identity matrix.

@@

In summary, the procedure for dimensional reduction is:
\begin{enumerate}
  \item Obtain the Fisher matrix $\matF$ of the system defined in
    \eqq{eq:fisher} using derivatives about the mean sample $\vecn_0,$
    plus the covariance matrix $\covm_n$ of the samples.
  \item From the eigensystem $(\matLam_n,\matV_n)$ of $C_n$, form the
    matrix $G$ defined in \eqq{eq:defG} and get its eigensystem
    $(\matLam_G,\matV_G).$  Place the eigenvalues in descending order.
  \item Choose the size $M$ of the compressed representation to be the
    minimum that keeps the $\left\langle\chi^2\right\rangle$ value in
    \eqq{eq:chiresid} below a chosen threshold, presumably $\ll 1.$
  \item The encoding matrix $E$ and decoding matrix $D$ are formed as
    in \eqq{eq:DE}, taking the inverse $\matLam_n^{-1/2}$ to be zero
    for any eigenvalues that are zero (or within roundoff errors).
  \item Compress all incoming (mean-subtracted) samples using \eqq{eq:uu}.  They will
    have unit covariance matrix and zero mean.
  \item Construct a continuous density estimator in $\vecu$ space that
    mimics the finite sample distribution.  If the distribution is
    normal, this becomes the multidimensional unit normal.  There
    is, however,  no \textit{a prior} reason that this must be the case,
    and something like a normalizing flow will be needed to
    approximate this lower-dimensional representation of the prior.
  \item Sample over $\vecu$ space in the Markov Chain that is sampling the
    posterior on the parameters of interest $\vecq,$ using \eqq{eq:UU}
    to transform each sample back into a $\hat\vecn$ vector.
  \end{enumerate}

@@

\section{Application}\label{sec:app}

xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx 
xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx 
xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx 
xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx

@@

>>>>

Our motivating situation is that we have a vector of observable summary statistics \vecc, such as the binned 2-point correlation functions of cosmic fields, that we are using to constrain a set of parameters of interest \vecq, which are cosmological parameters such as $\Omega_m, \sigma_8,$ etc.  There is a model $\hatc(\vecq,\vecn)$ for the observables which involves the parameters of interest, but also a vector \vecn\ of nuisance parameters, which include the coefficients of some linear expansion of the redshift distributions $n(z)$ of some population of the galaxies being observed:
\begin{equation}
  n(z) = \sum_{k=1}^{M} n_k b_k(z).
  \label{eq:nzbasis}
\end{equation}
The $b_k$ are a set of predetermined basis functions for the redshift distribution.  In our case there are multiple galaxy populations to be characterized, leading to hundreds of parameters $n_k$ to be considered.

We wish to characterized the Bayesian posterior probability
\begin{equation}
  p(\vecq | \vecc) \propto \int dn\, \likeli(\vecc | \vecq, \vecn) p(\vecq) p(\vecn),
\label{eq:posterior}
\end{equation}
where $\likeli(\vecc | \vecq, \vecn)$ is a known likelihood function
of the data, and $p(\vecq)$ and $p(\vecn)$ are priors on the
parameters.  This posterior is complex enough that it requires
approximation by the output of a Markov chain (MC) wandering across the space $(\vecq,\vecn).$

@@

For the DES Y3 analyses, \citet{hyperrank} devised a scheme whereby
the samples of \vecn\ are placed in a grid within some $M$-dimensional
unit hypercube $\mathcal{H}$.  The coordinates \vecu\ within the hypercube
are considered the compressed parameters of $n(z),$ and the
decompression function
$\hat{\vecn}(\vecu)$ outputs the $\vecn_\alpha$ sample at the nearest grid point to
  \vecu.  This solves the problem of creating a continuous \vecu\
  domain, and maintains the equal prior probability of each \vecn\
  sample, but the function \emph{output,} and the resultant likelihood
  function of \vecu, are discontinuous.
Various strategies are proposed to assign the $\vecn_\alpha$ to the grid points in
$\mathcal{H}$ based on summary statistics, to reduce the
discontinuities.  But the function is never smooth.
As a consequence, many MC samplers become quite inefficient in
sampling of the cosmological posterior.  In particular, samplers such as \textsc{MultiNest} that assume continuity are rendered nearly non-functional.  As a result, the Y3 cosmological priors could not be evaluated with this method.  Instead, the \vecn\ samples were not used, and an \textit{ad hoc} $p(\vecn)$ was adopted which allowed only shifts and dilations of the mean $n(z)$ of the \vecn\ samples.

@@

\section{Derivation}\label{sec:deriv}
We assume that we do have a multivariate normal likelihood for the
observables \vecc\, with the mean being some model
$\hatc(\vecq,\vecn)$ and a fixed covariance matrix $\covm_c.$ In this
section we will assume that the \vecn\ vectors have been shifted by
the mean $\vecn_0 \equiv \left\langle \vecn_\alpha \right\rangle$ so
that their mean is zero.
We seek some function $\hat{\vecn}(\vecu)$ of a lower-dimensional
vector \vecu\ which can be substituted for \vecn\ and yield nearly the
same likelihood function for any \vecn\ in the domain spanned by the
samples $\vecn_\alpha,\; \alpha\in 1\ldots N_\alpha.$
This means we want a map $\vecn_\alpha\rightarrow
\vecu_\alpha \rightarrow \hat{\vecn}_\alpha$ such that replacing $\vecn$ with
$\hat\vecn$ alters the cosmological inference by much less than the
other uncertainties.  We will implement this by minimizing the
distance in the space $\vecc$ between the
model generated by $\vecn$ and that by $\hat\vecn$, using the
observations' covariance matrix $\covm_c$ as a metric for the
distance.  This is equivalent to the $\chi^2$ of the difference
between the original and compressed models for the data:
\begin{equation} \left\langle \chi^2 \right\rangle
=  \frac{1}{N_\alpha} \sum_\alpha
                                            \left[ \hatc(\vecq,\vecn_\alpha) - \hatc(\vecq,\hat{\vecn}_\alpha) \right]^T
                                            \covm_c^{-1}
                                            \left[ \hatc(\vecq,\vecn_\alpha) - \hatc(\vecq,\hat{\vecn}\alpha) \right].
\label{eq:chisq}
\end{equation}
If the data are in fact drawn from the model $\hatc(\vecq,\vecn)$ with
a Gaussian likelihood, then this is also the mean shift in $-2\log\likeli$
from the compression.  It is \emph{not}, however, equal to the mean
shift of the overall log of the posterior in
\eqq{eq:posterior}---rather, $\left\langle\chi^2\right\rangle$ is serving as a
proxy for the true log-likelihood shift.

We next assume that the compression is linear, $\hat\vecn=\matX\vecn,$
for some matrix that is idempotent ($\matX\matX = \matX$).  If this is
true, then the contribution to \eqq{eq:chisq} at lowest order in
$\vecn-\hat\vecn$ is
\begin{equation}
  \left\langle \chi^2 \right\rangle = \frac{1}{N_\alpha} \sum_\alpha
  \left[ (\matI-\matX)\vecn_\alpha\right]^T \matF  \left[ (\matI-\matX)\vecn_\alpha\right]
  \label{eq:linearized}
\end{equation}
where we use the Jacobian matrix of the model $\hatc$ to define
\begin{equation}
  \matF \equiv
  \left[\frac{\partial\hatc}{\partial\vecn}\right]_{\vecq_0, \vecn_0}
  \covm_c^{-1} \left[\frac{\partial\hatc}{\partial\vecn}\right]_{\vecq_0,
    \vecn_0}^T.
\label{eq:fisher}
\end{equation}
This quantity is also the Fisher matrix over $\vecn$ giving the information
provided by the observations $\vecc.$  In many cases this matrix will be
rank-deficient and/or poorly conditioned, since the observables are not
likely to be very informative on $\vecn$---if they were, we might not be
concerned with establishing a prior on $\vecn$ to begin with.  We will
not, however, require the inverse of $\matF$ to proceed.

The optimization implied by \eqq{\eq:linearized} is the same as in
familiar Principal Components Analysis (PCA), aside from the presence
of the $F$ matrix, which in essence defines a new metric for the
variance to be captured by the principal components.  Our solution will follow the typical derivation
for PCA, but with an additional variable transformation to compensate
for the presence of $\matF.$

@@

Transforming the solution back into the space of $\vecn$ yields
\begin{align}
  \matX & = \matT^{-1} \matV_G \proj_M \matV_G^T \matT \\
\label{eq:DE}
   & = \left[ \matV_n^T \matLam_n^{1/2} \matV_G \proj_M \right] \left[
     \proj_M \matV_G^T \matLam_n^{-1/2} \matV_n^T \right] \\
        & \equiv \matD \matE.
\end{align}
We thus obtain our optimal encoding/compression using the nonzero rows of matrix
$\matE$ to give
\begin{equation}
  \vecu_\alpha = \matE \vecn_\alpha
  \label{eq:uu}
\end{equation}
and the decoding/reconstruction of the systematic variables as
\begin{equation}
  \hat\vecn_\alpha = \matD \vecu_\alpha.
  \label{eq:UU}
\end{equation}
One can confirm that this procedure yields a compressed representation
$\vecu$ such that $\covm_u = \matI_M,$ the $M$-dimensional
identity matrix.

@@

In summary, the procedure for dimensional reduction is:
\begin{enumerate}
  \item Obtain the Fisher matrix $\matF$ of the system defined in
    \eqq{eq:fisher} using derivatives about the mean sample $\vecn_0,$
    plus the covariance matrix $\covm_n$ of the samples.
  \item From the eigensystem $(\matLam_n,\matV_n)$ of $C_n$, form the
    matrix $G$ defined in \eqq{eq:defG} and get its eigensystem
    $(\matLam_G,\matV_G).$  Place the eigenvalues in descending order.
  \item Choose the size $M$ of the compressed representation to be the
    minimum that keeps the $\left\langle\chi^2\right\rangle$ value in
    \eqq{eq:chiresid} below a chosen threshold, presumably $\ll 1.$
  \item The encoding matrix $E$ and decoding matrix $D$ are formed as
    in \eqq{eq:DE}, taking the inverse $\matLam_n^{-1/2}$ to be zero
    for any eigenvalues that are zero (or within roundoff errors).
  \item Compress all incoming (mean-subtracted) samples using
    \eqq{eq:uu}.  The resulting $\vecu$ values will have
    have unit covariance matrix and zero mean.
  \item Construct a continuous density estimator in $\vecu$ space that
    mimics the finite sample distribution.  If the distribution is
    normal, this becomes the multidimensional unit normal.  There
    is, however,  no \textit{a priori} reason that this must be the case,
    and something like a normalizing flow may be needed to
    approximate this lower-dimensional representation of the prior.
  \item Sample over $\vecu$ space in the Markov Chain that is sampling the
    posterior on the parameters of interest $\vecq,$ using \eqq{eq:UU}
    to transform each sample back into a $\hat\vecn$ vector.
  \end{enumerate}

@@

\section{Application}\label{sec:app}
As an example of the application of this straightforward dimensional
reduction to a high-dimensional nuisance parameter, we examine the
redshift distribution of one of the bins of ``Maglim''  galaxies used
as a lens population and clustering tracer in the Year 6 analysis of
the \textit{Dark Energy Survey (DES)} galaxy catalogs.  In the baseline case,
there are 6 cosmological parameters of interest, $\vecq = \{\Omega_m,
\Omega_b, \sigma_8, h, n_s, m_\nu\}.$   The nuisance vector $\vecn$
has 5--10 parameters in each of the following categories: galaxy
biases with respect to matter; intrinsic alignments of galaxy shapes
with the tidal field of the mass; and multiplicative errors in the
measurement of galaxy shear.  The bulk of the nuisance parameters,
however, are the descriptions of the redshift distribution $n(z)$ of
galaxies belonging to each of 10 distinct color-based selections
applied to the galaxy catalog. Each galaxy sample's $n(z)$ distribution is
described by \eqq{eq:nzbasis} with 80 coefficients spanning $0<z<4$ at
intervals of $\Delta z=0.05,$  giving a total of 800 nuisance
parameters.  Each galaxy sample's $n(z)$ is also constrained to
integrate to unity, which yields 10 linear constraints on these
nuisance parameters.

Various forms of information from spectroscopic and photometric galaxy
surveys are used to constrain the $n(z)$ coefficients. As detailed in
\citet{y6pz}, these generate a set of samples of possible $n(z)$
coefficients for each galaxy sample.  We show here the results of the
mode compression process as applied to one of the nine galaxy samples,
the ??th lowest redshift of 6 lens bins.  We wish to have the
cosmological Markov chain sample from this
galaxy sample's 80-vector vector $\vecn$ of coefficients for its
redshift distribution, being guided by the available 3000??
equal-probability samples.

The observable quantities $\vecc$ whose modeling depends upon the these nuisance
parameters are the angular autocorrelation $w(\theta)$ of the sample
members, and the cross-correlations $\gamma_t^{(s)}(\theta)$ between
this sample's positions and the weak gravitational lensing shear
observed on the samples $1\le s \le 4$ of source galaxies.  A
covariance matrix spanning all of the experiment's observables is
available, and for this demonstration we will restrict it to the 30??
total elements of $\vecc$ that are affected by the $n(z)$ model for
this bin.

Figure showing violin plot, before/after.

Figure showing modes

Figure showing $\vecu$ distributions.

Describe 1d normalization mapping.

Figure showing histogram of chisq shifts.

@@

\bibiten[TBD]{y6pz} Boyan, William, ??? et al.\ 2025 (in preparation)
Y6 redshift paper(s).
